Two applications of logic programming to Coq

### 2020-11-05 papers/types2020/cultures.tex
Revised sections 1 and 2, responding to Alberto's comments.  Moving now to sections 3 and 5.

In many ways, proof theory is more elementary and low-level than most
approaches to type theory.  For example, type theories usually answer
the question ``What is a proof?'' with the response ``a (dependently)
typed $\lambda$-term''.  That is, when describing a type theory, one
usually decides that a proof is a certain kind of $\lambda$-term
within the system.  In contrast, the proof theory treats logical
propositions and proofs as separate.  For example, proof theory does
not assume that there are terms within the logic that describe proofs.

%% \begin{metanote}
%%   This paragraph seem to identify proof and type theory with sequents
%%   and nat ded and while I agree with the gist (although people such as
%%   Frank would argue that sequents are a way of organizing proof search
%%   for nat ded), we're not gonna make a lot of friends at types and
%%   more importantly is a big hammer to nail the point that logic
%%   programming is helpful with Coq --am
%% \end{metanote}

%% DM: I've tried to address your concerns by adding an additional
%% paragraph and references (below).

Gentzen's discovery that the key to treating classical and
intuitionistic logics in the same proof system was the invention of
the weakening and contraction structural rules on the right-side of
sequents~\cite{gentzen35}.  This invention lead him away from natural
deduction to multiple-conclusion sequent calculus, a very important
kind of proof that has not been given a satisfactory representation in
type theory.  This same innovation of Gentzen also opened the way to
another key proof-theoretic discovery, that of linear logic
\cite{girard87tcs}.  Similarly, sequent calculus provides an elegant
presentation of linear logic while most treatments of linear logic in
type theory have difficulty treating linear negation and the
multiplicative disjunction $\lpar$ and its unit $\bot$.

% DM I don't have a reference but I added more details.
%\begin{metanote}
%  `` partial fashion.'' Isn't that a bit vague? Some references? -am
%\end{metanote}

As is often observed, however, sequent calculus is too low-level to be
used explicitly as capturing the ``essence of a proof''.  Recent
developments in focused proof systems \cite{andreoli92jlc,liang09tcs},
makes it possible to extract from sequent calculus not only natural
deduction proofs \cite{pimentel16lsfa}, but also proof nets
\cite{chaudhuri08tcs}, and Herbrand-style expansion trees
\cite{chaudhuri16jlc}.  The role of focused proof systems to
characterize classes of proof structures is described in the next
section and used in our two example applications of Elpi with Coq.

%% DM I no longer think this comment is useful for this paper, so I
%% delete it instead of trying to fix it.

%% Proof theory also has rather immediate and natural treatments for
%% co-induction (as well as induction: see
%% \cite{baelde12tocl,momigliano12jal}) whereas the treatment of
%% coinduction within type theory remains more challenging and is well
%% developed only for various syntactic conditions (see, for example,
%% \cite{bertot08entcs}).
%% \begin{metanote}
%%   The criticism applies to Coq, less to type theory, see co-patterns --am
%% \end{metanote}

@@

Gentzen-style proofs are used to model computation in at least two
different ways.  The functional programming paradigm, following the
Curry-Howard correspondence, models computation abstractly as 
$\beta$-reduction of natural deduction proof: that is, computation is
modeled as \emph{proof normalization}.  On the other hand, the
logic programming paradigm, following the notion of goal-directed
proof search, models computation abstractly as a regimented search of
cut-free sequent calculus proofs: that is, computation is modeled
as \emph{proof search}.

Proof search has features that are hard to capture in the proof
normalization setting.  In particular, Gentzen's \emph{eigenvariables}
are a kind of a proof-level binder.  In the proof normalization
setting, such eigenvariables are instantiated during proof
normalization.  However, during the search for cut-free proofs,
eigenvariables do not vary, and they are part of the syntax of terms
and formulas.  As a result, they can be used in the treatment of
bindings in data structures more generally.  Such eigenvariables can
be used by the logic programming paradigm to provide a natural,
elegant, and powerful computing method with bindings within syntax.

@@

%% The following paragraph seems out-of-place so I delete it now.

%% In this paper, we shall use two different proof systems.  The first
%% logic is first-order intuitionistic logic.  Here, predicates
%% are not inductively defined and quantification will be limited to
%% range over first-order term.  Actually, we shall examine only the Horn
%% clause fragment of that logic.
%% %
%% The second logic will allow for inductively defined predicates using a
%% fixed point notation.   Given again our restriction to essentially
%% Horn clause specifications, it is possible to view intuitionistic
%% reasoning with such specification as if they are actually
%% specification in $\mu$MALL, an extension of multiplicative additive
%% linear logic with the least and greatest fixed point operators
%% \cite{baelde12tocl,baelde07lpar,heath19jar}.

@@

Given that \lP and Coq both result from a mixing of the
$\lambda$-calculus with logic, it is important to understand their
differences.  The confusion around the term \emph{higher-order
abstract syntax} (HOAS), for example, helps to illustrate their
differences.  In the functional programming setting, including the Coq
system, the HOAS approach leads to encoding binding structures within
terms using functions.  Such encodings are surprisingly complicated
since they allow for \emph{exotic terms} \cite{despeyroux95tlca} and
for structures on which induction is not possible
\cite{roeckl01fossacs}.  In the logic programming setting,
particularly \lP, HOAS is well supported since bindings are allowed
with terms ($\lambda$-bindings), in formulas (quantifiers), and in
proofs (eigenvariables).  For this reason, the term
\emph{$\lambda$-tree syntax} was introduced to name this particular
take on HOAS \cite{miller19jar}.  The Abella proof assistant
\cite{baelde14jfr} was designed, in part, to provide inductive and
co-inductive inference involving specifications using the
$\lambda$-tree syntax approach.

@@

While \lP is a typed language, it only uses simple, polymorphic types,
and these are more often used to indicate syntactic categories.  Any
dependency information needs to be captured using predicates and
quantifiers.

>>>>

In many ways, proof theory is more elementary and low-level than most
approaches to type theory.  For example, type theories usually answer
the question ``What is a proof?'' with the response ``a (dependently)
typed $\lambda$-term''.  That is, when describing a type theory, one
usually decides that a proof is a certain kind of $\lambda$-term
within the system.  In contrast, proof theory treats logical
propositions and proofs as separate.  For example, proof theory does
not assume that there are terms within the logic that describe proofs.

Gentzen's discovered that the key to treating classical and
intuitionistic logics in the same proof system was the weakening and
contraction structural rules on the right-side of
sequents~\cite{gentzen35}.  This discovery lead him away from natural
deduction to multiple-conclusion sequent calculus, a very important
kind of proof system that has not been given a satisfactory
representation in type theory.  This same innovation of Gentzen also
opened the way to another key proof-theoretic discovery, that of
linear logic \cite{girard87tcs}.  Similarly, sequent calculus provides
an elegant presentation of linear logic while most treatments of
linear logic in type theory have difficulty treating linear negation
and the multiplicative disjunction $\lpar$ and its unit $\bot$.

As is often observed, however, sequent calculus is too low-level to be
used explicitly as capturing the ``essence of a proof''.  Fortunately,
the notion of \emph{focused proof systems}
\cite{andreoli92jlc,liang09tcs}, makes it possible to collect and join
the micro-level inference rules of the sequent calculus into large
scale, synthetic inference rules.  As a result, using such proof
systems, it is possible to extract from sequent calculus not only
natural deduction proofs \cite{pimentel16lsfa}, but also proof nets
\cite{chaudhuri08tcs}, and Herbrand-style expansion trees
\cite{chaudhuri16jlc}.  The role of focused proof systems to
characterize classes of proof structures is described in the next
section and used in our two example applications of Elpi with Coq.

@@

Gentzen-style proofs are used to model computation in at least two
different ways.  The functional programming paradigm, following the
Curry-Howard correspondence, models computation abstractly as
$\beta$-reduction of natural deduction proof: that is, computation is
modeled as \emph{proof normalization}.  On the other hand, the logic
programming paradigm, following the notion of goal-directed proof
search, models computation abstractly as a regimented search for
cut-free sequent calculus proofs: that is, computation is modeled as
\emph{proof search}.

Proof search has features that are hard to capture in the proof
normalization setting.  In particular, Gentzen's \emph{eigenvariables}
are a kind of proof-level binder.  In the proof normalization setting,
such eigenvariables are instantiated during proof normalization.
However, during the search for cut-free proofs, eigenvariables do not
vary, and they are part of the syntax of terms and formulas.  As a
result, they can be used in the treatment of bindings in data
structures more generally.  Such eigenvariables can be used by the
logic programming paradigm to provide a natural, elegant, and powerful
computing approach to bindings within syntax.

@@

@@

Given that \lP and Coq both result from combining the
$\lambda$-calculus with logic, it is important to understand some of
their their differences.  The confusion around the term
\emph{higher-order abstract syntax} (HOAS), is a case in point.  In
the functional programming setting, including the Coq system, the HOAS
approach leads to encoding binding structures within terms using
functions.  The earliest such encodings were unsatisfactory since they
would allow for \emph{exotic terms} \cite{despeyroux95tlca} and for
structures on which induction was not possible \cite{roeckl01fossacs}.
Later approaches yielded non-canonical and complex encodings, although
these could support inductive and coinductive reasoning
\cite{chlipala08icfp,honsell01tcs,pientka10ijcar}. 
% DM I'm hoping that the additional sentence and references are enough
% here. 
%% \begin{metanote}
%%   AM: I find this a bit unbalanced. Exotic terms can be avoided with
%%   Honsell's theory of context and/or Adam's PHOAS -- not that I like
%%   it.. And of course there is Beluga and friends.  On the other hand
%%   Coq users are happy with library-based approaches such as Autosubt
%%   or well-typed DB that allows for sphisticated treatments of
%%   binders. I would say something more generic such as that mixing
%%   functions and HOAS is not immediate, may require additional
%%   machinery both in terms of type theories than of code generation,
%%   while in lp\dots
%% \end{metanote}
In the logic programming setting, particularly \lP, HOAS is well
supported since bindings are allowed with terms ($\lambda$-bindings),
in formulas (quantifiers), and in proofs (eigenvariables).  (In fact,
the original paper on HOAS \cite{pfenning88pldi} was inspired by \lP.)
For this reason, the term \emph{$\lambda$-tree syntax} was introduced
to name this particular take on HOAS \cite{miller19jar}.  The Abella
proof assistant \cite{baelde14jfr} was designed, in part, to provide
inductive and co-inductive inference involving specifications using
the $\lambda$-tree syntax approach.

@@

While \lP is a typed language, it only uses ML-style polymorphism
types, and these are more often used to indicate syntactic categories.
Any dependency information needs to be captured using predicates and
logical connectives and quantifiers.

% LocalWords:  Elpi polymorphically

### 2020-11-05 papers/types2020/intro.tex
Revised sections 1 and 2, responding to Alberto's comments.  Moving now to sections 3 and 5.

Recently, C. Sacerdoti Coen and E. Tassi developed the Elpi
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
more recently, Tassi has made Elpi available as the \texttt{coq-elpi} plugin to the Coq
proof assistant \cite{tassi18coqpl} (\url{https://github.com/LPCIC/coq-elpi}).  This implementation of \lP
extends earlier ones in primarily two directions:  First,
Elpi adds a notion of constraints and constraints handling rules, which makes it more expressive than
the Teyjus implementation \cite{nadathur99cade} of \lP. Second, the
plugin version of Elpi comes equipped with  a quotation and anti-quotation syntax for mixing
Coq expressions with \lP program elements and an API for accessing the
Coq environment, typechecking included.

@@

Since relations (not functions) are central in \lP, Elpi is
capable to providing direct support for the many relations that have a
role in proof assistants implementation and usage.  Such relations include typing (e.g.,
$\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural semantics
specifications \cite{kahn87stacs,hannan93jfp}), and interaction (e.g.,
structured operational semantics \cite{plotkin81,milner89book}).

Felty has also
made the point that LCF-style tactics and tacticals can be given an
elegant and natural specification using the higher-order relational
specifications provided by \lP \cite{felty93jar}.  Some recent
implementations built using Elpi
support the usefulness of
higher-order logic programming as a meta-programming language 
for proof assistants in general~\cite{DunchevCT16,coen19mscs} and in particular
for Coq via the \texttt{coq-elpi} plug in~\cite{tassi19itp,CohenST20} .
%

In this paper, we present two applications of Elpi within Coq. With
these examples, we shall illustrate that Elpi is useful not only
because of its meta-programming features,
% (e.g., a builtin treatment of bindings in syntax)
but also because it soundly
implements a higher-order intuitionistic logic.
\todo{Explain better? Our kernels are deep embedding, aren't they? -am}The latter aspect of
Elpi makes it possible to encode the proofs and proof theory of
specific subsets of the Coq logic.

While other meta-programming frameworks based on functional
programming such as \emph{MetaCoq}~\cite{sozeau2020metacoq} can and have been used for related endeavors, we believe (together with~\cite{DunchevCT16}) that they would require much more boilerplate.

>>>>

Recently, C. Sacerdoti Coen and E. Tassi developed the Elpi
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
more recently, Tassi has made Elpi available as the \texttt{coq-elpi}
plugin to the Coq proof assistant \cite{tassi18coqpl}
(\url{https://github.com/LPCIC/coq-elpi}).  This implementation of \lP
extends earlier ones in primarily two directions: First, Elpi adds a
notion of constraints and constraints handling rules, which makes it
more expressive than the Teyjus implementation \cite{nadathur99cade}
of \lP. Second, the plugin version of Elpi comes equipped with a
quotation and anti-quotation syntax for mixing Coq expressions with
\lP program elements and an API for accessing the Coq environment,
type checking included.

@@

Since relations (not functions) are central in \lP, Elpi is capable to
providing direct support for the many relations that have a role in
proof assistants implementation and usage.  Such relations include
typing (e.g., $\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural
semantics specifications \cite{kahn87stacs,hannan93jfp}), and
interaction (e.g., structured operational semantics
\cite{milner89book,plotkin81}).

Felty has also made the point that LCF-style tactics and tacticals can
be given an elegant and natural specification using the higher-order
relational specifications provided by \lP \cite{felty93jar}.  Some
recent implementations built using Elpi support the usefulness of
higher-order logic programming as a meta-programming language for
proof assistants in general~\cite{DunchevCT16,coen19mscs} and, in
particular, for Coq via the \texttt{coq-elpi} plug
in~\cite{CohenST20,tassi19itp} .

In this paper, we present two applications of Elpi within Coq. With
these examples, we shall illustrate that Elpi is useful not only
because of its meta-programming features, but also because it soundly
implements a higher-order intuitionistic logic: such implementations
of higher-order logic have long been known to provide powerful and
flexible approaches to implementing many different logics and their
proof systems \cite{felty89phd,felty88cade,paulson89jar}.  Following
that tradition, the Elpi system makes it possible to encode the proofs
and proof theory of various subsets of the logic behind Coq (see also
\cite{felty93lics}).

While other meta-programming frameworks based on functional
programming such as \emph{MetaCoq}~\cite{sozeau2020metacoq} can and
have been used for related endeavors, we believe (together
with~\cite{DunchevCT16}) that they would require much more
boilerplate code.
